# Introduction

**Learning objectives:**

- Review key challenges of statistical inference in general.

- Learn key challenges modelling of regression in particular.

## Three Challenges of Statistics

1. Generalizing from sample to population

2. Generalizing from treatment to control group 

3. Generalizing from observed measurements to the underlying constructs of interest 

All three challenges can  be framed as problems of prediction, and in this book we will focus on learning to address this with regression models.

## Why learn regression?

Regression is a method to summarize how predictions or averages of an *outcome* varies across individuals defined by a set of predictors.

Example: US Presidential elections vs economic growth in period leading up to the election:


```{r, echo=FALSE, message=FALSE}
library(rstanarm)
library(tidyverse)
```

```{r}
hibbs <- read_delim('data/hibbs.dat',trim_ws=TRUE, col_types='inncc')

```

```{r, message= FALSE, warning=FALSE, results='hide'}
M1 <- stan_glm(vote ~ growth, data=hibbs)
```

```{r}
ggplot(data=hibbs, aes(x=growth, y=vote)) +
  geom_point() + ylab('Incumbant Vote') + xlab('Precent Growth') +
  geom_abline(intercept = coef(M1)[1], slope = coef(M1)[2], col='blue')
```

We can examine the coefficients by printing the model:

```{r}
print(M1)
```

The median model is $y = 46.3 + 3.0x$ 

## Uses of a regresion model

- *Prediction*: For example, predict future vote (with some uncertainty!) given the economy.

- *Exploring Associations*: For example, the slope above summarizes the impact of the economy on election outcomes.

- *Extrapolation*: For example, pre-election polls can be adjusted for factors like party identification to extrapolate to the voting population at large. (Chapter 17.1)

- *Casual Inference*: Estimating treatment effects. For the election example, we can imagine looking at the effect some law (for example tax cuts) on election outcomes. (Chapter 19)

## Some examples

- *Estimating public opinion from an opt-in internet survey* - Challenge was extrapolation to general public from data collected from Xbox users. (Section 17.1)

- *Randomized experiment on the effect of an educational television program* - Challenge is adjusting for pre-treatment differences between treatment and control. (Section 19.2)

- *Estimating the effects of United Nations peacekeeping*  - Challenges include this being an observational study. A pre-treatment 'badness' score was used to correct for pre-treatment differences.

- *Estimating the effect of gun laws* - The challenge here was the difficulty of inference using regression with a
large number of predictors (50 data points with 30 predictors!!).  Also systematic differences between states were not captured by the models other predictors. 
 
We note that the peacekeeping study and the gun-control study are of the same time, but the peacekeeping study was more focused and was able to approximately account for systematic differences which does not seem possible with the gun-control study.

## Challenges with regression

Two main uses of regression

## Classical and Bayesian inference

## Computing least squares and Bayesian regression
 

## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/VYivNF3xL9c")`

<!--
### Cohort 2

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>
<summary> Meeting chat log </summary>

```
LOG
```
</details>
-->
