# Linear regression with a single predictor

**Learning objectives:**

- Introduce regression for a single predictor $y_i = a + b x_i + \text{error}$
- Demonstrate the steps of fitting, plotting and interpreting the fit. 
- Learn to check the procedure using fake-data simulation
- Learn how regression includes simple comparison as a simple case
 
## Example data  

```{r, message=FALSE}
library(ggplot2)
library(dplyr)
library(rstanarm)
```

This data comes from:

https://www.kaggle.com/datasets/spittman1248/cdc-data-nutrition-physical-activity-obesity

I will use a small subset, 2016 data containing for each state the percentage if adults who have obesity and the percent who engaged in strength training at least twice a week.

```{r}
exercise <- readr::read_csv('data/2016Exercise.csv',show_col_types = FALSE)
head(exercise)
```

```{r, echo=FALSE}
ggplot(exercise, aes(x= strengthTraining, y = obesity)) +  geom_text(aes(label= state))
```

## Fit the data {-}

```{r}
fit = stan_glm(obesity ~ strengthTraining, data = exercise, refresh= 0)
print(fit, detail=FALSE)
```
Note: `refresh=0` suppresses the usual stan chain progress reports.

Chapter 8 will discuss fitting procedure, but given the fit, what can we learn from this? 


## Examine the fit {-}

```{r, echo=FALSE}
ggplot(exercise) + geom_point( aes(x= strengthTraining, y = obesity))  +
                   geom_abline(aes(slope = coef(fit)[2], intercept = coef(fit)[1]))  + 
                   geom_text( aes(x= strengthTraining, y = obesity, label=state), hjust=1.2, vjust=1, alpha=0.6) +
                   geom_text( x = 24, y = 25, label = glue::glue('obesity = {round(coef(fit)[1],1)} - {-round(coef(fit)[2],1)} strengthTraining '))
```

- The fitted line is $y = 61.9 - 1.1 x$  The slope tells us that for every increase in percentage of people in a state that do strength training, we see a corresponding decrease in obesity by about 1.1 percent.   

- The uncertainties (MAD_SD) are given for the Intercept and the slope (`strengthTraining`), 3.9 and 0.1 respectively.

- The 95% interval for the slope is about -0.9 to -1.3, well separated from zero.

- The estimated residual standard deviation is 2.5: our linear model predicts obesity in a state within 2.5 percent points (68% range)

## Using the model to predict {-}

Suppose some state engages in a program to promote strength training, and manages over several years to successfully get a 32% participation rate. With this simple model, what would we predict for obesity rate in that state to be? 

```{r}
61.9 - 1.1*32
```
So we would predict 26.7% obesity, with an uncertainty of 2.5% points. (e.g. a 95% CI can be estimated as 21.7% to 31.7%) 

As the text states, this is understating the uncertainty since we have not included the uncertainty of the fitted parameters. This will be fixed in chapter 9

## Checking the procedure with fake-data simulation

- Step 1: Creating Pretend world.  

   Assume  $y = 61.9 - 1.1 x + \text{error}$ is true, $\text{error} \sim N(0,2.5)$ 
   
```{r}
a <- 61.9
b <- -1.1
sigma <- 2.5
## Use predictors from data set
x <- exercise$strengthTraining
n <- length(x)
```
   
- Step 2: Simulating fake data

```{r}
y <- a + b*x + rnorm(n,0,sigma)
fake <- tibble(x,y)
```
- Step 3: Fit the fake data

```{r}
fit <- stan_glm(y ~ x, data = fake, refresh = 0)
print(fit, detail=FALSE)
```
We can see in this particular simulation, the fit is consistent with the assumed model. But to understand this better we embed this in a loop...

## Fake Data Loop {-}

```{r, eval=FALSE}
n_fake <- 1000
cover_68 <- rep(NA, n_fake)
cover_95 <- rep(NA, n_fake)
for (s in 1:n_fake){
  y <- a + b*x + rnorm(n,0,sigma)
  fake <- tibble(x,y)
  fit <- stan_glm(y ~ x, data = fake, refresh = 0)
  b_hat <- coef(fit)["x"]
  b_se <- se(fit)["x"]
  # Is true value in the corresponding interval?
  cover_68[s] <- abs(b - b_hat) < b_se
  cover_95[s] <- abs(b - b_hat) < 2*b_se
}

cat(paste("68% coverage: ", mean(cover_68), "\n"))
cat(paste("95% coverage: ", mean(cover_95), "\n"))

```
```
68% coverage:  0.667 
95% coverage:  0.941
```

Not far from nominal values, which gives us confidence in the fitting procedure!

## Comparisons as Regression 

- Simple averages and comparisons can be interpreted as special cases of regression

- The key is to use an 'indicator' variable, a predictor that is 0 or 1 to `indicate` membership in a category. 

## Estimating the mean = regression on constant term {-}

```{r}
set.seed(42)
n_0 <- 20
y_0 <- rnorm(n_0, 2.0, 5.0)
cat(paste0('mean:', mean(y_0), '\nstandard error:', sd(y_0)/sqrt(n_0),"\n\n"))
fake_0 <- data.frame(y_0)
fit_0 <- stan_glm(y_0 ~ 1, data=fake_0,
prior_intercept=NULL, prior=NULL, prior_aux=NULL, refresh =0 )
print(fit_0, detail = FALSE)
```
Flat priors reproduce classical least squares estimate (more on this when we get to section 9.5)

What about default priors?

```{r}
fit_0 <- stan_glm(y_0 ~ 1, data=fake_0, refresh =0 )
print(fit_0, detail = FALSE)
```
## Estimating a difference = regressing on an indicator variable {-}

Add a new group:

```{r}
n_1 <- 30
y_1 <- rnorm(n_1, 8.0, 5.0)
diff <- mean(y_1) - mean(y_0)
se_0 <- sd(y_0)/sqrt(n_0)
se_1 <- sd(y_1)/sqrt(n_1)
se <- sqrt(se_0^2 + se_1^2)
cat(paste0("Diff: ",diff," Se: ", se))
```
Compare to true difference of 6.0

As a regression (again with flat priors):

```{r}
n <- n_0 + n_1
y <- c(y_0, y_1)
x <- c(rep(0, n_0), rep(1, n_1))
fake <- data.frame(x, y)
fit <- stan_glm(y ~ x, data=fake, prior_intercept=NULL, prior=NULL, prior_aux=NULL, refresh=0)
print(fit, detail=FALSE)
```
- Indicator slope (4.8) is the same as the difference in means

- MAD_SD (1.5) is nearly the same as the the SE

```{r, echo=FALSE}
ggplot(fake) + geom_point(mapping = aes(x=x, y=y)) +
               geom_abline(mapping = aes(slope = coef(fit)[2], intercept = coef(fit)[1])) +
               geom_text(x=0.5,y=7,label= "y = 2.9 + 4.8x") + 
               geom_hline(yintercept = mean(y_0), linetype = 'dashed') +
               geom_hline(yintercept = mean(y_1), linetype = 'dashed')
```


> Fake data simulation is a general tool that will continue to be helpful in more complicated settings.  

## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/hQFkChaqp1w")`

`r knitr::include_url("https://www.youtube.com/embed/wbV4eNkyNVU")`

<!--
### Cohort 2

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>
<summary> Meeting chat log </summary>

```
LOG
```
</details>
-->
